MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-06-20 latex/paper/sections/10-placement.tex
legalization

The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified. 
In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength (HPWL) of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.

@@

Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of our placer's SA algorithm. 
First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
Then, we initialize a simple geometric cooling schedule: 
\begin{equation}
    \label{coolingSchedule}
    T_{n+1} = \alpha T_n
\end{equation}
with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.

@@

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Movement approaches that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves.

>>>>

The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified. 
In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
We will be measuring wirelength using the manhattan distance between the 2-dimensional coordinates of the \texttt{SiteInst}s on the \texttt{device}.
We will refer to these distances as half perimeter wirelength (HPWL).
The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.

@@

Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of our placer's SA algorithm. 
First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
Then, we initialize a simple geometric cooling schedule: 
\begin{equation}
    \label{coolingSchedule}
    T_{n+1} = \alpha T_n
\end{equation}
with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
Hill-climbing moves can be accepted via the probability function \(P()\): 
\begin{equation}
    \label{moveAcceptance}
    P(s, s^*, T_n) = exp(\frac{E(s) - E(s^*)}{T_n})
\end{equation}
where \(E(s^*)\) is the energy of the proposed state \(s^*\), \(E(s)\) is the energy of the current state \(s\), and \(T_n\) is the current global temperature at \(n\) passes. 
If \(P(s, s^*, T_n) < random(0, 1)\), then \(s \leftarrow s^*\).
Recall that this probability is only evaluated for hill-climbing moves, thus the cost difference \(E(s) - E(s^*)\) is always negative.
In summary, a lower the cost difference and a higher the temperature contributes to a higher the probabilty of hill-climbing acceptance.
All hill-descending moves on the other hand are simply accepted outright.

@@

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators.

Directed moves as opposed to undirected moves give rise to the concept of legalization. 
When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection. 
However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.

In force directed moves, there is a concept of global placement versus detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
In this context, our placer follows a straightforward pack-legalize-place strategy.

### 2025-06-20 latex/paper/sections/11-placement_results.tex
legalization

Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five variants. 
\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590578.
This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600000 mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.

Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466855, which like the previous placer, lacks the diversity of random moves and contributes to faster crystallization into local minima.
However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation, which gives way to occasional hill-climbing. 
Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.

>>>>

Shown in figure \ref{fig:placers_overlay} are the total HPWL curves over number of passes for each of the five placers. 
\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.

Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466K.
Like the previous placer, \texttt{PlacerAnnealMidpoint} lacks the diversity of random moves which contributes to faster crystallization into local minima.
However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation which gives way to occasional hill-climbing. 
Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but can potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.

The next three placers performed significantly better than the previous two placers.
\texttt{PlacerAnnealRandom} and \texttt{PlacerAnnealHybrid} performed quite similarly with final costs of 346K and 322K respectively.
